---
title: "Project 3 Summary"
authors: "Ken Field, Riley McDonnell, Nellie Heitzman, Rob Han, Go Ogata, Will Snyder, Leila Hashemi, 
Alyssa Peeples, Fallon Goldberg, Brenna Prevelige, Savannah Weaver, and Alicia Kim"
date: "14 Apr 2020"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
if (!require("knitr")) install.packages("knitr"); library(knitr)
if (!require("UpSetR")) install.packages("UpSetR"); library(UpSetR)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse) # Load tidyverse last
```

## Project 3 Pipelines

| Name             | Github.Username  | Email.Address          | Pipeline          | Output    |
| :--------------- | :--------------- | :--------------------- | :---------------- | :-------- |
| Riley McDonnell  | rrm020           | <rrm020@bucknell.edu>  | kallisto-\>DESeq2 | Trans     |
| Nellie Heitzman  | Nellie001        | <nskh001@bucknell.edu> | Salmon-\>limma    | Both      |
| Owen LaFramboise | olaframboise     | <oll001@bucknell.edu>  | kallisto-\>edgeR  | Both      |
| Rob Han          | roberthjhan      | <rhh026@bucknell.edu>  | Sailfish-\>DESeq2 | Both      |
| Go Ogata         | go001            | <go001@bucknell.edu>   | STAR-\>limma      | Genes     |
| Will Snyder      | wsnyder4         | <wes021@bucknell.edu>  | STAR-\>edgeR      | Genes     |
| Leila Hashemi    | ly94-bot         | <lh046@bucknell.edu>   | Sailfish-\>limma  | Both      |
| Alyssa Peeples   | apeeples13       | <amp030@bucknell.edu>  | Salmon-\>DESeq2   | Both      |
| Fallon Goldberg  | fallongoldberg   | <feg003@bucknell.edu>  | kallisto-\>limma  | Trans     |
| Brenna Prevelige | brp006           | <brp006@bucknell.edu>  | Sailfish-\>edgeR  | Both      |
| Savannah Weaver  | science-with-sav | <sjw018@bucknell.edu>  | STAR-\>DESeq2     | Genes     |
| Alicia Kim       | ak039            | <ak039@bucknell.edu>   | Salmon-\>edgeR    | Both      |

```{r Loading Data, message = FALSE}
# Batch Comparison
batchfilelist <- list.files(path = "BatchResults", pattern = "*.csv", full.names = T)
batchfilenames <- list.files(path = "BatchResults", pattern = "*.csv", full.names = F)
batchfiles <- lapply(batchfilelist, read_csv)
batchfilenames <- gsub(".csv", "", batchfilenames)
names(batchfiles) <- batchfilenames
allbatchresults <- bind_rows(batchfiles, .id = "column_label")

# Genes
genefilelist <- list.files(path = "GeneResults", pattern = "*.csv", full.names = T)
genefilenames <- list.files(path = "GeneResults", pattern = "*.csv", full.names = F)
genefiles <- lapply(genefilelist, read_csv)
genefilenames <- gsub(".csv", "", genefilenames)
names(genefiles) <- genefilenames
allgeneresults <- bind_rows(genefiles, .id = "column_label")

# Transcripts
transcriptfilelist <- list.files(path = "TranscriptResults", pattern = "*.csv", full.names = T)
transcriptfilenames <- list.files(path = "TranscriptResults", pattern = "*.csv", full.names = F)
transcriptfiles <- lapply(transcriptfilelist, read_csv)
transcriptfilenames <- gsub(".csv", "", transcriptfilenames)
names(transcriptfiles) <- transcriptfilenames
alltranscriptresults <- bind_rows(transcriptfiles, .id = "column_label")
```

Note that `message = FALSE` was set to avoid outputting all of the column specifications.
Remove that option if you need to debug the column types.

### Gene Comparisons

```{r}
# Verify that each results table has expected padj and fold change cutoffs
allbatchresults %>%
  group_by(column_label) %>%
  summarise(max_padj = max(padj), min_foldchange = min(foldChange), max_foldchange = max(foldChange))

# Count the number of genes differentially expressed
allbatchresults %>%
  group_by(column_label) %>%
  dplyr::filter(str_detect(column_label, regex("DOWN", ignore_case = T))) %>%
  summarize(Down = length(ID)) -> batch_summary_down
allbatchresults %>%
  group_by(column_label) %>%
  dplyr::filter(str_detect(column_label, regex("UP", ignore_case = T))) %>%
  summarize(Up = length(ID)) -> batch_summary_up

#Combine
batch_summary_down$column_label <- gsub(".DOWN", "", batch_summary_down$column_label)
batch_summary_up$column_label <- gsub(".UP", "", batch_summary_up$column_label)
batch_summary <- merge(batch_summary_up, batch_summary_down, by = "column_label")
batch_summary <- rename(batch_summary, Pipeline = column_label)
batch_summary
```

### Gene Comparisons

Repeat the above summary for the transcript-level differential expression.

```{r}

```

### Transcript Comparisons

Repeat the above summary for the transcript-level differential expression.

```{r}

```

Which pipeline resulted in the greatest numbers of genes meeting the padj and fc cutoffs?

Does that necessarily mean that this pipeline was the "best"?

### Pipeline comparison

Compare the transcripts or genes identified in your pipeline to at least one other pipeline.

We could use a Venn diagram to make this comparison, but like pie charts these are not always the best visualization.

The UpSetR package includes better tools for visualizing the intersections of multiple sets. 
https://cran.rstudio.com/web/packages/UpSetR/vignettes/basic.usage.html

The following code plots the intersection between the Gene results for all of the files loaded. 

Use this example to compare your pipeline DE results to another pipeline.

```{r}
de_batch <- allbatchresults %>% 
  dplyr::select(Pipeline = column_label, ID)

intersection <- de_batch %>% mutate(value =1) %>% spread(Pipeline, value, fill =0 )
head(intersection)
intersection <- as.data.frame(intersection)
upset(intersection)
upset(intersection, empty.intersections = "on", order.by = "freq")
```


## Acknowledgements

Using lapply() to run functions over a list: https://astrostatistics.psu.edu/su07/R/html/base/html/lapply.html

Converting a list into a single data frame: https://stackoverflow.com/questions/2851327/convert-a-list-of-data-frames-into-one-data-frame

UpSetR: https://cran.rstudio.com/web/packages/UpSetR/

Converting gene lists into UpSetR format: https://crazyhottommy.blogspot.com/2016/01/upset-plot-for-overlapping-chip-seq.html

Upset plots for genome analysis: https://genomespot.blogspot.com/2017/09/upset-plots-as-replacement-to-venn.html 