---
title: "ideal Lab"
author: "Ken Field"
date: "1/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ideal")
library(ideal)
if (!require("airway")) BiocManager::install("airway"); library(airway)
if (!require("org.Hs.eg.db")) BiocManager::install("org.Hs.eg.db"); library(org.Hs.eg.db)
if (!require("DESeq2")) install.packages("DESeq2"); library(DESeq2)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

## ideal Tutorial

The dataset contains four cell lines in two conditions: control and treatment with dexamethasone.

Himes BE, Jiang X, Wagner P, Hu R, Wang Q, Klanderman B, Whitaker RM, Duan Q, Lasky-Su J, Nikolos C, Jester W, Johnson M, Panettieri R Jr, Tantisira KG, Weiss ST, Lu Q. 'RNA-Seq Transcriptome Profiling Identifies CRISPLD2 as a Glucocorticoid Responsive Gene that Modulates Cytokine Function in Airway Smooth Muscle Cells.' PLoS One. 2014 Jun 13;9(6):e99625. PMID: 24926665. GEO: GSE52778.

```{r Tutorial}
data(airway)
airway
dds_airway <- DESeq2::DESeqDataSetFromMatrix(assay(airway),
                                             colData = colData(airway),
                                             design=~cell+dex)
# run deseq on it
dds_airway <- DESeq(dds_airway)
# extract the results
res_airway <- results(dds_airway, contrast = c("dex","trt","untrt"),alpha = 0.05)
## Launching ideal - Choose one of the following options
ideal(dds_airway)
# providing the results object
ideal(dds_obj = dds_airway,res_obj = res_airway)
# providing the annotations
genenames_airway <- mapIds(org.Hs.eg.db,keys = rownames(dds_airway),column = "SYMBOL",keytype="ENSEMBL")
annotation_airway <- data.frame(gene_id = rownames(dds_airway),
                                gene_name = genenames_airway,
                                row.names = rownames(dds_airway),
                                stringsAsFactors = FALSE)
head(annotation_airway)  
ideal(dds_obj = dds_airway, res_obj = res_airway,
      annotation_obj = annotation_airway)
```

It might be best to run the first option and then walk through the rest interactively.